Predicting the Capture Rate of Pokemon

### 2022-12-12 Predicting_Pokemon_Capture_Rate.Rmd
finishing up

---
title: "Predicting the Capture Rate of Pokemon"
subtitle: "PSTAT 131 Final Project"
author: "Shivani Kharva"
date: "2022-12-11"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

@@

@@

@@

@@

```{r, message = FALSE, warning = FALSE}
# Loading all necessary packages
library(tidyverse)
library(dplyr)
library(tidymodels)
library(readr)
library(kknn)
library(janitor)
library(ISLR)
library(discrim)
library(poissonreg)
library(glmnet)
library(corrr)
library(corrplot)
library(randomForest)
library(xgboost)
library(rpart.plot)
library(vip)
library(ranger)
library(tidytext)

@@

@@

@@

@@

@@

```{r}
# Distribution of capture rate
ggplot(pokemon_data, aes(capture_rate)) +
  geom_bar(fill='red3')
```

The possible values of `capture_rate` range from 0 to 255 and the `capture_rate` of the Pokemon in our data appears to range from 3 to 255. There is also a peak at about 45, meaning the most common capture rate for a Pokemon in our data is 45.

@@

The data contains 7 different generations of Pokemon. The generations contain Pokemon from each of the 7 generations of Pokemon games. This poses the question of whether there are specific generations that have higher capture rates overall than any other generations.

```{r}
# Boxplot of generation vs. capture rate
ggplot(pokemon, aes(x=generation, y=capture_rate, group=generation)) +
  geom_boxplot(aes(fill = generation))
```

@@

### Legendary Status

Legendary Pokemon are especially powerful Pokemon that players rarely encounter. From this definition, it seems likely that it would be harder to capture a Pokemon that is legendary than one that is not. Let's see if this is true or not.

```{r}
ggplot(pokemon_data, aes(capture_rate)) + 
  geom_bar(aes(fill = is_legendary)) +
  scale_fill_manual(values = c('red2', 'blue4'))
```

From the graph, we can see that, although there are still non-legendary Pokemon in the lower end of the capture rate, there are no legendary Pokemon with a capture rate above 50. Also, the lowest capture rate in our data, 3, is made up of almost all legendary Pokemon. It seems that legendary Pokemon have a lower capture rate than non-legendary Pokemon, which is what was expected.

@@

# Visualization of correlation matrix
pokemon_corrplot <- corrplot(pokemon_cor, method = "number")
```

One relationship that stands out is the relatively strong negative correlation between `base_total` and `capture_rate` (-0.71), which indicates that it may be useful to use `base_total` in our analysis rather than the six individual battle statistics (we cannot use all of them because the battle statistics add up to the exact `base_total`). This relationship implies that, as a Pokemon has a higher total of battle statistics, it becomes more difficult to capture them (lower `capture_rate`).

@@

@@

We will use k-fold stratified cross validation in order to train the data on 10 mini datasets (or folds) created using the training data. Since we will be essentially using 10 mini training/testing sets to train our data, this will give us a better estimate of each model's true performance. We stratify on the outcome, `capture_rate`, to make sure the data in each fold is not imbalanced.

@@

It is now time to build our models! Since the models take a very long time to run, the results from each of the models has been saved to avoid rerunning the models every time. I have chosen Root Mean Squared Error (RMSE) as my metric because it works as an overall metric for all models. The RMSE is one of the most commonly used measures for evaluating the performance of regression models by showing how far the model's predictions are from the true values using Euclidian distance. Since RMSE measures distance, it is important that we normalize our data, which we did in the recipe. Also, I have fit 8 models to the Pokemon data; however, we will only be conducting further analysis on the 4 best-performing models. Let's get to building our models!

@@

```{r, eval=FALSE, warning=FALSE, message=FALSE}
# LINEAR REGRESSION 
lm_model <- linear_reg() %>% 
  set_engine("lm")

@@

```{r, eval=FALSE, warning=FALSE, message=FALSE}
# LINEAR REGRESSION 
lm_workflow <- workflow() %>% 
  add_model(lm_model) %>% 
  add_recipe(pokemon_recipe)

@@

# LASSO REGRESSION
lasso_workflow <- workflow() %>% 
  add_recipe(ridge_lasso_recipe) %>% 
  add_model(lasso_spec)

@@

```{r, eval=FALSE, warning=FALSE, message=FALSE}
# collect_metrics() to collect the RMSE
# slice() to save only RMSE

@@

# Creating a barplot of the RMSE values
ggplot(all_models, aes(x=Model, y=RMSE)) +
  geom_bar(stat = "identity", aes(fill = Model)) +
  scale_fill_manual(values = c("blue1", "red1", "blue2", "red2", "blue3", "red3", "blue4", "red4")) +
  theme(legend.position = "none") 
```
*Note: the colors in this bar plot do not represent anything and are simply for aesthetic preference.*

@@

@@

For the boosted trees, we tuned the learning rate, number of trees, and minimal node size with 5 different levels. It may be useful to note that some tweaking of the model was done prior to finalizing the ranges for these parameters. The model did substantially worse when the minimal node size was a lower number, which is why the range 40 to 60 was chosen. Also, the model did not do as well when the learning rate was too high. A high learning rate causes the model to learn faster, but it also trains the model less and makes it less generalized. So, the range 0.01 to 0.1 was chosen to see which smaller learning rate value performs the best. From the starting points of each of the plots, it appears that the model does better at a higher learning rate. This means that the model does better when it is learning faster. Also, it appears that generally, once there are more than ~50 trees, the number of trees does not make much of a change on the performance of the model. However, the model appears to do worse when there are less than ~50 trees. The minimal node size does not seem to have much of an effect on the performance of the model.

@@

@@

```{r}
# Creating plot of predicted values vs. actual values
pokemon_tibble %>% 
  ggplot(aes(x = .pred, y = capture_rate)) +
  geom_point(alpha = 0.4) +
  geom_abline(lty = 2) +
  theme_grey() +
  coord_obs_pred()
```

@@

```{r}
# Using the training fit to create the VIP because the model was not actually fit to the testing data
rf_final_fit_train %>% 
  extract_fit_engine() %>% 
  vip(aesthetics = list(fill = "red3", color = "blue3"))
```

@@

>>>>

---
title: "Predicting the Capture Rate of Pokemon"
subtitle: "PSTAT 131 Final Project"
author: "Shivani Kharva"
date: "2022-12-11"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    df_print: paged
    theme: journal
---

@@

<center>![](https://media.tenor.com/8vuqpD3Ir-IAAAAC/catching-pokemon.gif)</center>

@@

<center>![Pokemon in battle!](https://static.wikia.nocookie.net/essentialsdocs/images/7/70/Battle.png/revision/latest?cb=20220523172438)</center>

@@

<center>![Pokemon stats screen.](https://78.media.tumblr.com/76016a9930e876e104b528190af0e125/tumblr_oxbo3x64Ge1w5miaho1_1280.png)</center>

@@

```{r, message = FALSE, warning = FALSE}
# Loading all necessary packages
library(tidyverse)
library(dplyr)
library(tidymodels)
library(readr)
library(kknn)
library(janitor)
library(ISLR)
library(discrim)
library(poissonreg)
library(glmnet)
library(corrr)
library(corrplot)
library(randomForest)
library(xgboost)
library(rpart.plot)
library(vip)
library(ranger)
library(tidytext)
library(ggplot2)
theme_set(theme_bw())

@@

<center>![Left: Minior with its shell (Meteorite). Right: Minior without its shell (Core).](https://i1.wp.com/hallowpeakgaming.com/wp-content/uploads/2016/12/minior.jpg?fit=1024%2C486&ssl=1)</center>

@@

<center>![](https://images-wixmp-ed30a86b8c4ca887773594c2.wixmp.com/f/2aaacf79-3286-4339-9843-f60dfc484cc3/dety9on-74a91eb7-7f6a-43e7-a87f-67bcb6a4796b.gif?token=eyJ0eXAiOiJKV1QiLCJhbGciOiJIUzI1NiJ9.eyJzdWIiOiJ1cm46YXBwOjdlMGQxODg5ODIyNjQzNzNhNWYwZDQxNWVhMGQyNmUwIiwiaXNzIjoidXJuOmFwcDo3ZTBkMTg4OTgyMjY0MzczYTVmMGQ0MTVlYTBkMjZlMCIsIm9iaiI6W1t7InBhdGgiOiJcL2ZcLzJhYWFjZjc5LTMyODYtNDMzOS05ODQzLWY2MGRmYzQ4NGNjM1wvZGV0eTlvbi03NGE5MWViNy03ZjZhLTQzZTctYTg3Zi02N2JjYjZhNDc5NmIuZ2lmIn1dXSwiYXVkIjpbInVybjpzZXJ2aWNlOmZpbGUuZG93bmxvYWQiXX0.rFMKqBaSDLqE8knqyrMLA-DOqL7KOLTl7iawmzojIyI)</center>

@@

<center>![](https://media.tenor.com/tEGO5Q-EQIcAAAAC/pokemon.gif)</center>

@@

Now, we will explore the relationships between select variables with the outcome as well as with each other!

@@

```{r}
# Distribution of capture rate
ggplot(pokemon_data, aes(capture_rate)) +
  geom_bar(fill='red3') +
  labs(
    title = "Distribution of Capture Rate"
  )
```

The possible values of `capture_rate` range from 0 to 255 and the `capture_rate` of the Pokemon in our data appears to range from 3 to 255. There is also a peak at about 45, meaning the most common capture rate for a Pokemon in our data is ~45.

@@

The data contains 7 different generations of Pokemon. The generations contain Pokemon from each of the 7 generations of Pokemon games. This poses the question of whether there are specific generations that have higher capture rates overall than any other generations. Also, legendary Pokemon are especially powerful Pokemon that players rarely encounter. From this definition, it seems likely that it would be harder to capture a Pokemon that is legendary than one that is not. Let's check to see how these variables relate to one another and to the capture rate!

```{r}
# Boxplot of generation and legendary status vs. capture rate
pokemon_data %>% 
  ggplot(aes(x = capture_rate, y = reorder(is_legendary, capture_rate), fill = generation)) +
  geom_boxplot() +
  labs(
    title = "Box Plot of Generation and Legendary vs. Capture Rate",
    y = "Legendary Status", 
    x = "Capture Rate") +
  scale_fill_manual(values = c("red1", "blue1", "red3", "blue3", "darkred", "blue4", "#660000"))
```

@@

It is also apparent that there are very few Pokemon, if any, of legendary status in any generations other than 6 and 7. However, the distribution of generations 6 and 7 across legendary Pokemon still fit the aforementioned observations.

As for legendary status, from the plot, we can see that, although there are still non-legendary Pokemon in the lower end of the capture rate, there are no legendary Pokemon with a capture rate above 50. All legendary Pokemon capture rates appears to be less than all first quartiles of non-legendary Pokemon capture rates. The data implies that legendary Pokemon genearally have a lower capture rate than non-legendary Pokemon, which is what was expected.

### Type 1 Plot

In our analysis, we are only using Type 1 of each Pokemon since not all Pokemon have a Type 2. However, how many types of Pokemon are there?

```{r}
# Creating a bar chart of `type1`
pokemon_data %>% 
  ggplot(aes(y = type1)) +
  geom_bar(fill = "#003399") +
  labs(
    title = "Count of Each Type"
  )
```

From the plot, we can see that there are 18 different types for type 1 Pokemon. 18 types is a lot of categories to include! However, the plot also reveals that there are significantly more Pokemon in some types like "water" and "normal" in comparison to other types with much fewer Pokemon like "flying" and "fairy". So, in our analysis we should specify that we only want to prioritize the types with the most Pokemon and group the rest of the types into "Other". This will be completed and further discussed in the recipe creation.

@@

# Visualization of correlation matrix
pokemon_corrplot <- corrplot(pokemon_cor, method = "circle", addCoef.col = 1, number.cex = 0.7)
```

One relationship that stands out is the moderately negative correlation between `base_total` and `capture_rate` (-0.71), which indicates that it may be useful to use `base_total` in our analysis rather than the six individual battle statistics (we cannot use all of them because the battle statistics add up to the exact `base_total`). This relationship implies that, as a Pokemon has a higher total of battle statistics, it becomes more difficult to capture them (lower `capture_rate`).

@@

### Base Total

Since we discussed the relationship between `base_total` and `capture_rate` in the correlation plot, perhaps we should further explore the relationship between the two variables.

```{r, message = FALSE, warning=FALSE}
pokemon_data %>% 
  ggplot(aes(x=base_total, y=capture_rate)) + 
  geom_jitter(width = 0.5, size = 1) +
  geom_smooth(method = "lm", se =F, col="darkred") +
  labs(title = "Capture Rate vs. Base Total")
```

Although the relationship is not very strong, it is apparent that there is a moderately negative relationship between `capture_rate` and `base_total`. That is, a higher value of total battle statistics slightly relates to a lower capture rate. This makes contextual sense because a Pokemon with greater battle abilities would be more valuable (and harder to capture because of its greater value). So, it seems as if it would prove more meaningful to use `base_total` rather than the six battle statistics in exploring the relationship between the predictors and the outcome, `capture_rate`.

@@

We will create 10 folds to conduct k-fold (10-fold in our case) stratified cross validation. This means that R is taking the training data and assigning each observation in the training data to 1 of 10 folds. For each fold, a testing set is created consisting of that fold and the remaining k-1 folds will be the training set for that fold. At the end, we end up with k total folds.

K-fold cross validation is done by splitting the data into k folds as described above with each fold being a testing set with the other k-1 folds being the training set for that fold. Then, whichever model we are fitting is fit to each training set and tested on the corresponding testing set (each time, a different fold should be used as a validation set). Then, the average accuracy is taken from the testing set of each of the folds to measure performance (other metrics can be taken as well such as standard error).

We use k-fold cross validation rather than simply fitting and testing models on the entire training set because cross validation provides a better estimate of the testing accuracy. It is better to take the mean accuracy from several samples instead of just one accuracy from one sample because, as n increases, we reduce variation.

We stratify on the outcome, `capture_rate`, to make sure the data in each fold is not imbalanced.

@@

It is now time to build our models! Since the models take a very long time to run, the results from each of the models has been saved to avoid rerunning the models every time. I have chosen Root Mean Squared Error (RMSE) as my metric because it works as an overall metric for all models. The RMSE is one of the most commonly used measures for evaluating the performance of regression models by showing how far the model's predictions are from the true values using Euclidian distance. So, a lower RMSE is better since that means the predicted values have a smaller distance from the actual values. Since RMSE measures distance, it is important that we normalize our data, which we did in the recipe. Also, I have fit 8 models to the Pokemon data; however, we will only be conducting further analysis on the 4 best-performing models. Let's get to building our models!

<center>![](https://pbs.twimg.com/media/EYoS4TmXgAIfLRD.jpg)</center>

@@

```{r, warning=FALSE, message=FALSE}
# LINEAR REGRESSION 
lm_model <- linear_reg() %>% 
  set_engine("lm")

@@

```{r, warning=FALSE, message=FALSE}
# LINEAR REGRESSION 
lm_workflow <- workflow() %>% 
  add_model(lm_model) %>% 
  add_recipe(pokemon_recipe)

@@

# LASSO REGRESSION
lasso_workflow <- workflow() %>% 
  add_recipe(pokemon_recipe) %>% 
  add_model(lasso_spec)

@@

```{r, warning=FALSE, message=FALSE}
# collect_metrics() to collect the RMSE
# slice() to save only RMSE

@@

# Creating a barplot of the RMSE values
ggplot(all_models, aes(x=Model, y=RMSE)) +
  geom_bar(stat = "identity", aes(fill = Model)) +
  scale_fill_manual(values = c("blue1", "red1", "blue2", "red2", "blue3", "red3", "blue4", "red4")) +
  theme(legend.position = "none") +
  labs(title = "Comparing RMSE by Model")
```

*Note: the colors in this bar plot do not represent anything and are simply for aesthetic preference.*

@@

<center>![](https://media4.giphy.com/media/6nWhy3ulBL7GSCvKw6/giphy.gif)</center>

@@

For the boosted trees, we tuned the learning rate, number of trees, and minimal node size with 5 different levels. It may be useful to note that some tweaking of the model was done prior to finalizing the ranges for these parameters. The model did substantially worse when the minimal node size was a lower number, which is why the range 40 to 60 was chosen. Also, the model did not do as well when the learning rate was too high. A high learning rate causes the model to learn faster, but it also trains the model less and makes it less generalized. So, the range 0.01 to 0.1 was chosen to see which smaller learning rate value performs the best. From the starting points of each of the plots, it appears that the model does better at a higher learning rate. This means that the model does better when it is learning faster. Also, it appears that generally, once there are more than \~50 trees, the number of trees does not make much of a change on the performance of the model. However, the model appears to do worse when there are less than \~50 trees. The minimal node size does not seem to have much of an effect on the performance of the model.

@@

<center>![](https://www.ungeek.ph/wp-content/uploads/2016/11/ash_sparkiling-eyes.gif)</center>

@@

```{r}
# Creating plot of predicted values vs. actual values
pokemon_tibble %>% 
  ggplot(aes(x = .pred, y = capture_rate)) +
  geom_point(alpha = 0.4) +
  geom_abline(lty = 2) +
  theme_grey() +
  coord_obs_pred() +
  labs(title = "Predicted Values vs. Actual Values")
```

@@

```{r}
# Loading in the training data fit
rf_final_fit_train <- read_rds(file = "data/rf_final/rf_final_train.rds")

# Using the training fit to create the VIP because the model was not actually fit to the testing data
rf_final_fit_train %>% 
  extract_fit_engine() %>% 
  vip(aesthetics = list(fill = "red3", color = "blue3"))
```

@@

<center>![](https://i.pinimg.com/originals/8f/23/e4/8f23e48ae48e3a07b342a718249e0080.gif)</center>
